On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-05-31 Manuscript/manuscript.tex
Major changes before Tuesday meeting

For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
throughout the paper. 
%\of{Ok I updated this.  Should I delete the batch variable too?  Although I don't
%use it, I do reference it in the methods section}
The pooled data increases the sample size which is advantageous because in some
years there are very few final heat observations.  For example, in 2022 there
were only five data points due to two disqualifications and one athlete who 
did not compete.

@@

Although it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other. 
\eds{Did you try it and find that the effect was not needed?  If yes,
I would include that here.  That argument 
would be more compelling for its non-use than its difficulty in interpretation.}

@@

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the pseudo-sample.
The test statistic $S$ is the average of $W^*$ averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a statistic $Z$ which follows standard normal distribution
asymptotically \citep[p.910]{datta2005rank}. An implementation of this method is
available from the \texttt{clusWilcox.test()} function with
\texttt{method = `ds'} from R package \texttt{clusrank}
\citep{jiang2017wilcoxon}.
\eds{are you including preliminary heat data as possible within cluster 
observations in this analysis?  Based on your 
discussion in Section 2.1 about how they are not comparable, it may not be fair 
to include them here?  Or perhaps do the analysis with and without prelim heat
data to see if it makes a difference?}

@@

\of{Add something like this to the start of the paragraph?}
We may want to consider using all four data sets for this problem because
comparing all the data to the data excluding 2022 can give us a qualitative
assessment of how impactful 2022 was.
\eds{Agree with JY, but you need to describe what these four datasets are.}

Table~\ref{tab:Gamma_parameters} summarizes the results from the gamma model 
from different data sets.

As shown in Table~\ref{tab:Gamma_parameters}, the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022.  The men's final data venue effect error increased from 0.439
to 0.504 and from 0.0495 to 0.0554 for the men's pooled data.  This demonstrates
that the inclusion of 2022 increased the randomness of the data and motivates
the rank-based study (see Section~\ref{subsec:Results_Rank}).

\jy{Close captions with a period.}
\begin{table}
  \centering
  \caption{Summary statistics for the GLMM on various data sets. 
	\eds{Int=intercept?}}
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Older Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Older Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

\eds{be consistent with the order of the rows between 
Tables~\ref{tab:Gamma_parameters} and~\ref{tab:Sim_probability}}

\jy{label the table and reference to it; clean it up}
\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than 0.1, 0.09, and
  0.08 for various data sets.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & P(RxnTime $<$ 0.1) & P(RxnTime $<$ 0.09) & P(RxnTime $<$ 0.08) \\ 
   \midrule
   all finals & $8.56\cdot10^{-4}$ & $4.1\cdot10^{-5}$ & $6.0\cdot10^{-7}$ \\ 
   old finals & $4.39\cdot10^{-4}$ & $1.63\cdot10^{-5}$ & $2.0\cdot10^{-5}$ \\
   all pooled & $6.86\cdot10^{-3}$ & $1.21\cdot10^{-3}$ & $1.3\cdot10^{-4}$ \\
   old pooled & $5.76\cdot10^{-3}$ & $9.93\cdot10^{-4}$ & $1.0\cdot10^{-4}$\\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

Using the gamma function with link set to log, the probability of the reaction 
time being less than 0.1 seconds can be calculated to see if Allen's 
disqualification was an anomaly. \eds{explain how simulation was used here.}
At a size of $n=10,000,000$ \eds{use different letter as $n$ was already used 
for number of athletes, $B?$}, the calculated probabilities of a reaction time 
being below 0.1 seconds for the various different data sets are shown in 
Table~\ref{tab:Sim_probability}. Prior to the 2022 World Championships, there was 
a 0.0439\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals.
That means that we would expect a time below the threshold to occur
once every 2,277 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (the World Championships are biennial), every 569 years there
should be a time as extreme as Allen's. This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

>>>>

For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
throughout the paper. 
%\of{Ok I updated this.  Should I delete the batch variable too?  Although I don't
%use it, I do reference it in the methods section}
The pooled data increases the sample size which is advantageous because in some
years there are very few final heat observations.  For example, in 2022 there
were only five data points due to two disqualifications and one athlete who 
did not compete.  We also look at data that does not include 2022 with year values
between 1999 and 2019.  Thus, we have four datasets with their respective naming
in parentheses: the final's data from 1999 to 2019 (Old Finals), final's data 
from 2019 to 2022 (All Finals), pooled data from 1999 to 2019 (Old Pooled), and 
pooled data from 1999 to 2022 (All Pooled).

@@

Although it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other. 
\eds{Did you try it and find that the effect was not needed?  If yes,
I would include that here.  That argument 
would be more compelling for its non-use than its difficulty in interpretation.}
\of{We found that including the race effect resulted in the venue effect not
being significant, and thus Prof Yan suggested that we don't include it.  I am
wondering if this is even worth mentioning? Prof Yan, what do you think?}

@@

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the pseudo-sample.
The test statistic $S$ is the average of $W^*$ averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a statistic $Z$ which follows standard normal distribution
asymptotically \citep[p.910]{datta2005rank}. An implementation of this method is
available from the \texttt{clusWilcox.test()} function with
\texttt{method = `ds'} from R package \texttt{clusrank}
\citep{jiang2017wilcoxon}.
\eds{are you including preliminary heat data as possible within cluster 
observations in this analysis?  Based on your 
discussion in Section 2.1 about how they are not comparable, it may not be fair 
to include them here?  Or perhaps do the analysis with and without prelim heat
data to see if it makes a difference?}
\of{Yes I am including prelminary heat data.  The primary reason is that there
would be few datapoints otherwise.  I ran the numbes without prelims, the number
of observations dropped from 65 to 26 and the number of clusters from 15 to 9, the
p value increased from 0.000155 to 0.007332.
The calculations are at the bottom of RankComparison.Rmd}

@@

\of{Add something like this to the start of the paragraph?}
We may want to consider using all four data sets for this problem because
comparing all the data to the data excluding 2022 can give us an
assessment of how impactful 2022 was. We hope to see that by including 2022,
the probability of observing a significantly low reaction time increases. 
\eds{Agree with JY, but you need to describe what these four datasets are.}
\of{I added a description of the four datasets in the Data paragraph right below
where it says <Thus, we have four datasets>}
Table~\ref{tab:Gamma_parameters} summarizes the results from the gamma model 
from different data sets and shows the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022.  The men's final data venue effect error increased from 0.439
to 0.504 and from 0.0495 to 0.0554 for the men's pooled data.  This demonstrates
that the inclusion of 2022 increased the randomness of the data and motivates
the rank-based study (see Section~\ref{subsec:Results_Rank}).

\jy{Close captions with a period.}
\begin{table}
  \centering
  \caption{Summary statistics for the GLMM on various data sets.} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Old Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Old Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

\jy{label the table and reference to it; clean it up}
\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than 0.1, 0.08, and
  0.08 for various data sets.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & P(RxnTime $<$ 0.1) & P(RxnTime $<$ 0.08) \\ 
   \midrule
   Old Finals & $4.39\cdot10^{-4}$ & $2.0\cdot10^{-5}$ \\
   All Finals & $8.56\cdot10^{-4}$ & $6.0\cdot10^{-7}$ \\
   Old Pooled & $5.76\cdot10^{-3}$ & $1.0\cdot10^{-4}$ \\ 
   All Pooled & $6.86\cdot10^{-3}$ & $1.3\cdot10^{-4}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

\begin{table}
  \centering
  \caption{Suggested reaction time barriers based on probability of observing
  a time less the barrier. \of{Not sure if this makes sense}}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & P(RxnTime $<$ x = .001) & P(RxnTime $<$ x = 0.0001) \\ 
   \midrule
   Old Finals & $4.39\cdot10^{-4}$ & $2.0\cdot10^{-5}$ \\
   All Finals & $8.56\cdot10^{-4}$ & $6.0\cdot10^{-7}$ \\
   Old Pooled & $5.76\cdot10^{-3}$ & $1.0\cdot10^{-4}$ \\ 
   All Pooled & $6.86\cdot10^{-3}$ & $1.3\cdot10^{-4}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:ExpectedValues}
\end{table}

Using the gamma function with link set to log, the probability of the reaction 
time being less than 0.1 seconds can be calculated to see if Allen's 
disqualification was an anomaly. 
\eds{explain how simulation was used here.} \of{The methods section already
details how the calculations are performed, can you please elaborate what you
are exactly looking for.}
Calculations were performed in R using the methods described in 
section~\ref{sec:glmm} with the use of the \citet{lme4} package.

At a size of $n=10,000,000$, \eds{use different letter as $n$ was already used 
for number of athletes, $B?$} \of{Could I just say "At a size of 10,000,000"? 
Otherwise I am fine with using "B" instead of "n"}

The calculated probabilities of a reaction time 
being below 0.1 seconds for the various different data sets are shown in 
Table~\ref{tab:Sim_probability}. Prior to the 2022 World Championships, there was 
a 0.0439\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals.
That means that we would expect a time below the threshold to occur
once every 2,277 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (the World Championships are biennial), every 569 years there
should be a time as extreme as Allen's. This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.
